Detecting Treatment-Subgroup Interactions in Clustered Data with Generalized Linear Mixed-Effects Model Trees

### 2017-05-23 pkg/glmertree/paper/treatment_interactions.Rnw
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1370 255f4249-2263-4417-b1ea-8498779e27a7

\section{Abstract}
Identification of subgroups of patients for which treatment A is more effective than treatment B, and vice versa, is of key importance to the development of personalized medicine. Tree-based algorithms are helpful tools for the  detection of such interactions, but none of the available algorithms allow for taking into account clustered or nested dataset structures, which are particularly common in psychological research. Therefore, we propose the generalized linear mixed-effects model tree (GLMM tree) algorithm, which allows for the detection of treatment-subgroup interactions, while accounting for the clustered structure of a dataset. The algorithm uses model-based recursive partitioning to detect treatment-subgroup interactions, and a GLMM to estimate the random-effects parameters. In a simulation study, GLMM trees show higher accuracy in recovering treatment-subgroup interactions, higher predictive accuracy, and lower Type-II error rates than GLM-based recursive partitioning and mixed-effects regression trees. Also, GLMM trees show somewhat higher predictive accuracy than linear mixed-effects models with pre-specified interaction effect, on average. We illustrate the application of GLMM trees on an individual patient-level data meta-analysis on treatments for depression. We conclude that GLMM trees are a promising exploratory tool for the detection of treatment-subgroup interactions in clustered datasets.\\
\\
\textit{Keywords}: model-based recursive partitioning, treatment-subgroup interactions, random effects, generalized linear mixed-effects model, classification and regression trees

@@

Due to the recursive nature of the splitting, the rectangular regions of the partition can be graphically depicted as nodes in a decision tree, as shown in the artificial example in Figure~\ref{fig:exampletree}. The partition in Figure~\ref{fig:exampletree} is rather simple, based on the values of two predictor variables: duration and anxiety. The resulting tree has a depth of two, as the longest path travels along two splits. Each of the splits in the tree is defined by a splitting variable and value. The first split in the tree separates the observations into two subgroups, based on the duration variable and a splitting value of 8, yielding two rectangular regions, represented by node~2 and node~3. As the observations in node~2 are not further split, node~2 is a terminal node. Node~3 is an inner node, as the observations in this node are further split into nodes~4 and~5, based on the anxiety variable.

@@

We will use an artificial motivating dataset with treatment-subgroup interactions to introduce the GLMM tree algorithm. This dataset consists of a set of observations on $N = 150$ patients, who were randomly assigned to one of two treatment alternatives (Treatment~1 or Treatment~2). The treatment outcome is represented by the variable depression, quantifying post-treatment depressive symptomatology. The potential moderator variables are duration, age and anxiety. Duration reflects the number of months the patient has been suffering from depression prior to treatment, age reflects patients' age in years at the start of treatment and anxiety reflects patients' total scores on an anxiety inventory administered before treatment. Summary statistics of these variables are provided in Table~\ref{tab:sum_stats}. Each patient was part of one of ten clusters, each having a different value for the random intercept, which was generated from a standard normal distribution and uncorrelated with the partitioning variables.

@@

\begin{description}
  \item[Step 0:] Initialize by setting $r$ and all values $\hat{b}_{(r)}$ to 0.

@@

  \item[Step 3:] Repeat Steps~1 and~2 until convergence.
\end{description}

@@

To assess the performance of GLMM trees, we carried out three simulation studies: In Study~I we assessed and compared the accuracy of GLMM trees, linear-model based MOB and mixed-effects regression trees (MERTs) in datasets with treatment-subgroup interactions. In Study~II, we assessed and compared the Type-I error of GLMM trees and linear-model based MOB in datasets without treatment-subgroup interactions. In Study~III, we assessed and compared the performance of GLMM trees and linear mixed-effects models (LMMs) with pre-specified interactions in datasets with piecewise and continuous interactions. As the outcome variable was continuous in all simulated datasets, the GLMM tree algorithm and trees resulting from its application will be referred to as LMM tree(s).

@@

R \cite{R} was used for data generation and analyses. The \verb|partykit| package (version 1.0-2; \citeNP{HothyZeil15}) was employed for estimating LM trees, using the \verb|lmtree| function. For estimation of LMM trees the \verb|lmertree| function of the \verb|glmertree| package (version 0.1-0; \citeNP{FokkyZeil15}; available from R-Forge) was used. The significance level $\alpha$ for the parameter instability tests was set to .05 for all trees, with a Bonferroni correction applied for multiple testing. The minimum number of observations per node in trees was set to 20 and maximum tree depth was set to three, thus limiting the number of terminal nodes to eight in every tree.

The \verb|REEMtree| package \cite{SelaySimo11} was employed for estimating MERTs, using default settings. For estimating LMMs the \verb|lmer| function function from the \verb|lme4| package (version 1.1-7; \citeNP{BateyMeac12}) was employed, using restricted maximum likelihood (REML) estimation. The \verb|lmerTest| package (version 2.0-32; \citeNP{KuznyBroc16}) was used to assess statistical significance of fixed-effects predictors in LMMs in Study~III. The \verb|lmerTest| package calculates effective degrees of freedom and $p$-values based on Satterthwaite approximations.

@@

In this simulation design, some of the potential partitioning covariates are true partitioning covariates, the others are noise variable. Therefore, an extra level was added to the sixth facet of the \textit{General simulation design}:

@@

An additional facet was added to assess the effect of the magnitude of treatment-effect differences:

@@

Predictive accuracy of each method was assessed by calculating the correlation between true and predicted treatment-effect differences. To prevent overly optimistic estimates of predictive accuracy \cite{HastyTibs09}, predictive accuracy was assessed using test datasets. Test datasets were generated from the same population as training datasets, but test observations were not drawn from the same clusters as the training observations, but from `new' clusters.

For MERT, predicted treatment-effect differences were obtained by fitting two MERTs on the training data: one using observations in the first treatment condition and one using observations in the second treatment condition. Predictions of treatment-effect differences for test observations were obtained by dropping test observations down both trees and taking the difference between the two predicted values. This approach was taken as it yielded higher predictive accuracy than the alternative of fitting a single MERT using treatment as a potential partitioning variable.

@@

The average size of LMM trees was 7.15 nodes ($\mathrm{SD}=0.61$), whereas the average size of LM trees was 8.15 nodes ($\mathrm{SD}=2.05$), indicating that LM trees tend to involve more spurious splits than LMM trees. The effects of the most important predictors of tree size are depicted in Figure~\ref{fig:xyplot_treesize}. The average size of LMM trees was close to the true tree size in all conditions. In the absence of random effects, this was also the case for LM trees. In the presence of random effects that are correlated to a (potential) partitioning variable, LM trees start to create spurious splits, especially with larger $\sigma_b$ values. In the presence of random effects that are uncorrelated to the other variables in the model, LM trees lack power to detect treatment-subgroup interactions if sample size is small (i.e., $N=200$). With larger sample sizes, LM trees showed about the true tree size, on average.

@@

Further splits were more accurately recovered by LMM trees yielding 90.40\% accuracy for the full partition comparted to only 61.44\% for LM trees. The effects of the four most important predictors of tree accuracy are depicted in Figure~\ref{fig:xyplot_treeaccuracy}. In the absence of random effects, LM and LMM trees were about equally accurate. In the presence of random effects, LM trees were much less accurate than LMM trees when random effects were correlated with a partitioning covariate. When random intercepts were not correlated with one of the $U_k$ variables, LMM trees outperformed LM trees only when sample size was small (i.e., $N = 200$).

@@

The predicted treatment-effect differences of LMM trees show an average correlation of .93 (SD = .13) with the true differences. LM trees and MERTs show lower accuracy, with an average correlations of .88 ($\mathrm{SD}=.19$) and .75 ($\mathrm{SD}=.21$), respectively. The most important predictors of predictive accuracy are depicted in Figure~\ref{fig:xyplot_correlations}. Performance of all three algorithms improves with increasing sample size and treatment-effect differences. Furthermore, LMM trees and MERTs are not much affected by the presence and magnitude of random effects in the data. LMM trees perform most accurately in most conditions and are never outperformed by the other methods. MERTs perform the least accurate in most conditions and never outperforms the other methods, but the differences in accuracy become less pronounced with larger sample and effect sizes.

@@

To assess the effect of the treatment-effect difference $\beta$, an additional facet was added to the \textit{General simulation design}: 
\begin{enumerate}
\item[7.] Two levels for $\beta$, the global mean difference in treatment outcomes: $\beta = 2.5$ (corresponding to a medium effect size, Cohen's $d = 0.5$) and $\beta = 5.0$ (corresponding to a large effect size; Cohen's $d = 1.0$). 
\end{enumerate}

@@

The interactions in Study~I (Figure~\ref{fig:modelC}) can be referred to as piecewise interactions, as their effect is a stepwise function of the moderator (partitioning) variables. Trees are preeminently suited for recovering such piecewise or subgroup interactions, but may have difficulty when the true interactions are continuous functions of moderator variables (for example, $U_1 \cdot U_2$). At the same time, linear regression models with pre-specified interaction terms may perform well in recovering continuous interactions, but may have difficulty in recovering piecewise interactions. Therefore, in the third simulation study, we added a seventh facet to the \textit{General simulation design} described above:

@@

For the datasets with both piecewise and continuous interactions, the partition as depicted in Figure~\ref{fig:modelC} was also used. In addition, the fixed-effects part $x_{i}^{\top}\beta_{j}$ in each of the terminal nodes now comprised continuous main and (treatment) interaction effects of the partitioning variables. In other words, the partitioning variables $U_2$, $U_1$ and $U_5$ appear are also part of the linear predictor $x_i$, as part of the terms presented in Table~\ref{tab:continuous_terms}. The corresponding node-specific $\beta_j$ parameters are also presented in Table~\ref{tab:continuous_terms}. The $\beta_j$ values were chosen to yield the same treatment-subgroup means as in Figure~\ref{fig:modelC}. The interaction terms were created using centered $U_k$ variables, calculated by subtracting their variable means. Again, the outcome variable was calculated as $y_i = x_{i}^{\top}\beta_{j} + z_{i}^{\top}b + \epsilon_i$, with the value of $\beta_j$ depending on the values of $U_2$, $U_1$ and $U_5$.

In datasets with purely continuous interactions, $\beta$ has a global value and no subscript, comprising only purely continuous main and interaction effects, as shown by the single $\beta$ column in Table~\ref{tab:continuous_terms}. The outcome variable was calculated as $y_i = x_{i}^{\top}\beta + z_{i}^{\top}b + \epsilon_i$.

@@

As the outcome variable is continuousTo this end, the MOB algorithm tests for parameter stability with respect to
each of a set of auxiliary covariates, we employed an identity link and Gaussian response distribution. The resulting GLMM trees will therefore be refered to as LMM trees. To compare the accuracy of LMM trees, we also fitted LM trees and LMMs with pre-specified interactions to the data. In the LMMs, the outcome variable was regressed on a random intercept, main effects of treatment and the potential moderators (partitioning variables) and interactions between treatment and the potential moderators. As it is not known in advance how to interact the potential moderators, higher-order interactions were not included.

@@

Figure~\ref{fig:lmtree_C&W} and Figure~\ref{fig:lmertree_C&W} present the LM and LMM trees fitted on the IPDMA data. The LM tree (Figure~\ref{fig:lmtree_C&W}) selected level of education as the first partitioning variable, and presence of a comorbid anxiety disorder as a second partitioning variable, for observations with a higher level of education. By taking into account study-specific intercepts, the LMM tree (Figure~\ref{fig:lmertree_C&W}) indicates that the first split in the LM tree may be spurious. The LMM tree selected presence of a comorbid anxiety disorder as the only partitioning variable. The terminal nodes of Figure~\ref{fig:lmertree_C&W} show only a single treatment-subgroup interaction: for patients without a comorbid anxiety disorder, CBT and PHA provide more or less the same reduction in HAM-D scores (Cohen's $d = 0.05$). For patients with a comorbid anxiety disorder, PHA provides a greater reduction in HAM-D scores (Cohen's $d = 0.39$). The estimated intraclass correlation coefficient for the GLMM tree was .05.

@@

The LMM with pre-specified treatment interactions yielded three significant predictors of treatment outcome: like in the GLMM tree, an effect of the presence of a comorbid anxiety disorder was found (main effect: $b = 2.29$, $p = .002$; interaction with treatment: $b = -2.10$, $p = .028$). Also, the GLMM indicated an interaction between treatment and age ($b = .10$, $p = .018$).

@@

We presented the GLMM tree algorithm, which allows for estimation of a GLM-based recursive partition, as well as estimation of random-effects parameters. We hypothesized GLMM trees to be well suited for the detection of treatment-subgroup interactions in clustered datasets and confirmed this by our simulation studies.

GLMM trees accurately recovered the subgroups in 90\% of simulated datasets with treatment-subgroup interactions, outperforming GLM trees without random effects. In terms of predictive accuracy, GLMM trees outperformed GLM trees as well as MERTs, predicting treatment-effect differences in test data with .94 accuracy, on average. In datasets without treatment-subgroup effects, GLMM trees showed a Type-I error rate of 4\%, compared with a Type-I error rate of 33\% for GLM trees.

Better performance of GLMM trees was mostly observed when random effects were sizeable and correlated with potential partitioning variables. In those circumstances, GLM trees are likely to detect spurious splits and subgroups. At the same time, GLM trees showed less power to detect subgroups in the presence of random effects. As expected, accuracy of MERTs was not affected by the presence of random effects, but only approached GLMM trees' accuracy in datasets with the largest sample and effect sizes. Especially with smaller effect and sample sizes (i.e., Cohen's $d=.5$ and/or $N = 200$), GLMM tree outperformed the other tree methods. Such effect and sample sizes are quite common in multi-center clinical trials, and GLMM trees may provide a useful tool for subgroup detection in those instances.

In the absence of random effects, GLM and GLMM trees yielded very similar predictive accuracy. This finding is of practical importance, indicating that application of GLMM trees will not `hurt' predictive accuray when in fact, random effects are absent.

Compared to linear mixed-effects models with pre-specified interactions, GLMM trees provided somewhat better accuracy on average. GLMM trees performed poorly in datasets with purely continuous interactions, but much better than GLMMs when interactions were at least partly piecewise. GLMM trees performed better in the presence of larger numbers of potential moderator variables, indicating that GLMM trees are much better suited for exploration than GLMMs.

In the Application, we obtained similar results: the GLMM tree yielded somewhat higher predictive accuracy, while using a smaller number of variables than the other two methods (GLM tree and a GLMM with pre-specified interactions). The GLMM trees obtained over repeated subsamples of the training data proved to be relatively stable.

\subsubsection{Limitations}

\readme{I would emphasize early on that GLMM trees rely on the properties of their building blocks, namely GLMMs and MOBs. The effect of misspecifications are studied to some degree for both methods. For GLMM one could cite papers discussing this. For MOB, or trees more broadly, there are also some references. One of these could be the `To Split or to Mix?' paper by Frick, Strobl, and Zeileis.}

Recursive partitioning methods were origfinally developed as a non-parametric tool for classification and regression, assuming the mechanism that generated the data unknown \cite<e.g., >{Brei01TwoCult}. GLMM trees take MOB and GLMMs as building blocks, thereby relying on the properties of these building blocks. As a result, GLMM trees may inherit their sensitivities to model misspecification as well as their robustness against it. We will address potential types of misspecification in fitting GLMM trees:

\readme{MF: GLMM tree at least tackles one type of random-effects misspecification: ignoring them. \cite{BarryLevy13} and \cite{Barr13} advice to keep it maximal is at repeated measures designs. \cite{McCuyNeuh11}: "A key conclusion is the large degree of robustness of maximum likelihood for a wide variety of commonly encountered situation". .... fit fixed-effects linear models in the nodes of the tree and a global model for the random effects, in turn introducing several distributional assumptions about the random effects and errors. Misspecification of these distributions will likely have a negative effect on the accuracy of the estimated GLMM trees.}

Furthermore, misspecification of partitioning and fixed- and random-effects variables will also reduce accuracy of the resulting GLMM trees. If relevant variables are omitted or incorrectly specified, GLMM trees can only approximate the true subgroups using the specified variables. Our simulations indicate that LM trees detect spurious subgroups as a result of misspecifying (that is, not including) the random effects. Reduced accuracy and spurious splits can also be expected to occur when relevant random effects are not included when specifying the GLMM tree model. Furthermore, as the random effects are estimated globally, misspecification of the random effects may have a strong impact on the resulting GLMM tree model.

Another source of misspecification is the inclusion of irrelevant variables. Although our simulations indicate that the performance of GLMM trees was not negatively affected by increasing the number of noise variables specified for partitioning from 2 to 12, the power to detect subgroups may be reduced with much larger numbers of noise variables. Including irrelevant variables in the random or fixed effects may also negatively affect accuracy of GLMM trees, but we have not assessed this in our study.

Users can evaluate potential misspecification by inspecting residuals and predictions, and assessing tree stability. The vigette included as Supplementary Material shows how residuals andrandom-effects predictions can be plotted to assess potential misspecification. In the Application we have shown how tree stability can be assessed using the methods suggested by \cite{PhilyZeil16}. Future studies may focus on the effects of possible misspecification on the performance of GLMM trees.

As GLMM trees fit more complex models than non-parametric tree-based methods, like CART for example, larger sample sizes are needed to fit the model. How much larger likely depends on the complexity of the specified model: Our simulations show that with fixed and random-effects specifications with only a single predictor, a sample size of 200 is sufficient to detect subgroups with moderate differences in treatment effect. More complex fixed- and random-effects specifications will require larger sample sizes; how these affect the performance of GLMM trees requires further research.

\subsubsection{Future directions}

\readme{MF: Add something about implementing non-normal non-parametric or mixture ranef}

In the Introduction we mentioned several existing tree-based methods for treatment-subgroup interaction detection. These methods have different objectives and there is not yet an agreed-upon single best method. In a simulation study, \citeA{SiesyVanM16} found the method of \citeA{ZhanyTsia12Stat} to perform best, followed by MOB. However, the method of Zhang et al. performed worst under some conditions of the simulation study in terms of the Type I error rate. Further research comparing tree-based methods for treatment-subgroup interaction detection is needed, especially  for clustered datasets, as our simulations and comparisons only focused on GLMM trees and GLM trees.

>>>>

\section{Abstract}
Identification of subgroups of patients for whom treatment A is more effective than treatment B, and vice versa, is of key importance to the development of personalized medicine. Tree-based algorithms are helpful tools for the  detection of such interactions, but none of the available algorithms allow for taking into account clustered or nested dataset structures, which are particularly common in psychological research. Therefore, we propose the generalized linear mixed-effects model tree (GLMM tree) algorithm, which allows for the detection of treatment-subgroup interactions, while accounting for the clustered structure of a dataset. The algorithm uses model-based recursive partitioning to detect treatment-subgroup interactions, and a GLMM to estimate the random-effects parameters. In a simulation study, GLMM trees show higher accuracy in recovering treatment-subgroup interactions, higher predictive accuracy, and lower Type-II error rates than linear-model based recursive partitioning and mixed-effects regression trees. Also, GLMM trees show somewhat higher predictive accuracy than linear mixed-effects models with pre-specified interaction effects, on average. We illustrate the application of GLMM trees on an individual patient-level data meta-analysis on treatments for depression. We conclude that GLMM trees are a promising exploratory tool for the detection of treatment-subgroup interactions in clustered datasets.\\
\\
\textit{Keywords}: model-based recursive partitioning, treatment-subgroup interactions, random effects, generalized linear mixed-effects model, classification and regression trees

@@

Due to the recursive nature of the splitting, the rectangular regions of a recursive partition can be graphically depicted as nodes in a decision tree, as shown in the artificial example in Figure~\ref{fig:exampletree}. The partition in Figure~\ref{fig:exampletree} is rather simple, based on the values of two predictor variables: duration and anxiety. The resulting tree has a depth of two, as the longest path travels along two splits. Each of the splits in the tree is defined by a splitting variable and value. The first split in the tree separates the observations into two subgroups, based on the duration variable and a splitting value of 8, yielding two rectangular regions, represented by node~2 and node~3. As the observations in node~2 are not further split, node~2 is a terminal node. Node~3 is an inner node, as the observations in this node are further split into nodes~4 and~5, based on the anxiety variable.

@@

We will use an artificial motivating dataset with treatment-subgroup interactions to introduce the GLMM tree algorithm. This dataset consists of a set of observations on $N = 150$ patients, who were randomly assigned to one of two treatment alternatives (Treatment~1 or Treatment~2). The treatment outcome is represented by the variable depression, quantifying post-treatment depressive symptomatology. The potential moderator variables are duration, age and anxiety. Duration reflects the number of months the patient has been suffering from depression prior to treatment, age reflects patients' age in years at the start of treatment and anxiety reflects patients' total scores on an anxiety inventory administered before treatment. Summary statistics of these variables are provided in Table~\ref{tab:sum_stats}. Each patient was part of one of ten clusters, each having a different value for the random intercept, which were generated from a standard normal distribution and uncorrelated with the partitioning variables.

@@

\begin{minipage}{\linewidth}
\begin{description}
  \item[Step 0:] Initialize by setting $r$ and all values $\hat{b}_{(r)}$ to 0.

@@

  \item[Step 3:] Repeat Steps~1 and~2 until convergence.

\end{description}
\end{minipage}
\\
\\

@@

To assess the performance of GLMM trees, we carried out three simulation studies: In Study~I we assessed and compared the accuracy of GLMM trees, linear-model based MOB (LM trees) and mixed-effects regression trees (MERTs) in datasets with treatment-subgroup interactions. In Study~II, we assessed and compared the Type-I error of GLMM trees and linear-model based MOB in datasets without treatment-subgroup interactions. In Study~III, we assessed and compared the performance of GLMM trees and linear mixed-effects models (LMMs) with pre-specified interactions in datasets with piecewise and continuous interactions. As the outcome variable was continuous in all simulated datasets, the GLMM tree algorithm and trees resulting from its application will be referred to as LMM tree(s).

@@

R \cite{R} was used for data generation and analyses. The \verb|partykit| package (version 1.0-2; \citeNP{HothyZeil15, pkg:partykit}) was employed for estimating LM trees, using the \verb|lmtree| function. For estimation of LMM trees the \verb|lmertree| function of the \verb|glmertree| package (version 0.1-0; \citeNP{FokkyZeil15}; available from R-Forge) was used. The significance level $\alpha$ for the parameter instability tests was set to .05 for all trees, with a Bonferroni correction applied for multiple testing. The minimum number of observations per node in trees was set to 20 and maximum tree depth was set to three, thus limiting the number of terminal nodes to eight in every tree.

The \verb|REEMtree| package (version 0.9.3; \citeNP{SelaySimo11}) was employed for estimating MERTs, using default settings. For estimating LMMs the \verb|lmer| function function from the \verb|lme4| package (version 1.1-7; \citeNP{BateyMeac15, pkg:lme4}) was employed, using restricted maximum likelihood (REML) estimation. The \verb|lmerTest| package (version 2.0-32; \citeNP{KuznyBroc16}) was used to assess statistical significance of fixed-effects predictors in LMMs in Study~III. The \verb|lmerTest| package calculates effective degrees of freedom and $p$-values based on Satterthwaite approximations.

@@

In this simulation design, some of the potential partitioning covariates are true partitioning covariates, the others are noise variable. Therefore, in addition to the \textit{General simulation design}, the following facet was added in the this study:

@@

To assess the effect of the magnitude of treatment-effect differences, the following facet was added in this study:

@@

Predictive accuracy of each method was assessed by calculating the correlation between true and predicted treatment-effect differences. To prevent overly optimistic estimates of predictive accuracy, predictive accuracy was assessed using test datasets. Test datasets were generated from the same population as training datasets, but test observations were not drawn from the same clusters as the training observations, but from `new' clusters.

The best approach for including treatment effects in MERTs is not completely obvious. Firstly, a single MERT may be fitted, where treatment is included as one of the potential partitioning variables. Predictions of treatment-effect differences can then be obtained by dropping test observations down the tree twice, once for every level of the treatment indicator. Secondly,  two MERTs may be fitted: one using observations in the first treatment condition and one using observations in the second treatment condition. Predictions of treatment-effect differences can then be obtained by dropping a test observation down each of the two trees. We tried both approaches: the second approach yielded higher predictive accuracy, as the first approach often did not pick up the treatment indicator as a predictor. Therefore, we have taken the second approach of fitting two MERTs to each dataset in our simulations.

@@

The average size of LMM trees was 7.15 nodes ($\mathrm{SD}=0.61$), whereas the average size of LM trees was 8.15 nodes ($\mathrm{SD}=2.05$), indicating that LM trees tend to involve more spurious splits than LMM trees. The effects of the most important predictors of tree size are depicted in Figure~\ref{fig:xyplot_treesize}. The average size of LMM trees was close to the true tree size in all conditions. In the absence of random effects, this was also the case for LM trees. In the presence of random effects that are correlated to a (potential) partitioning variable, LM trees start to create spurious splits, especially with larger $\sigma_b$ values. In the presence of random effects that are uncorrelated to the other variables in the model, LM trees lack power to detect treatment-subgroup interactions if sample size is small (i.e., $N=200$). With larger sample sizes, LM trees showed about the true tree size, on average. Tree size of MERTs was not assessed, as a single true tree size for MERTs could not be derived from the design in Figure~\ref{fig:modelC}.

@@

Further splits were more accurately recovered by LMM trees yielding 90.40\% accuracy for the full partition comparted to only 61.44\% for LM trees. The effects of the four most important predictors of tree accuracy are depicted in Figure~\ref{fig:xyplot_treeaccuracy}. In the absence of random effects, LM and LMM trees were about equally accurate. In the presence of random effects, LM trees were much less accurate than LMM trees when random effects were correlated with a partitioning covariate. When random intercepts were not correlated with one of the $U_k$ variables, LMM trees outperformed LM trees only when sample size was small (i.e., $N = 200$). Tree accuracy of MERTs was not assessed, as a single accurate tree structure for MERTs could not be derived from the design in Figure~\ref{fig:modelC}.

@@

The predicted treatment-effect differences of LMM trees show an average correlation of .93 ($\mathrm{SD = .13}$) with the true differences. LM trees and MERTs show lower accuracy, with an average correlations of .88 ($\mathrm{SD}=.19$) and .75 ($\mathrm{SD}=.21$), respectively. The most important predictors of predictive accuracy are depicted in Figure~\ref{fig:xyplot_correlations}. Performance of all three algorithms improves with increasing sample size and treatment-effect differences. Furthermore, LMM trees and MERTs are not much affected by the presence and magnitude of random effects in the data. LMM trees perform most accurately in most conditions and are never outperformed by the other methods. MERTs perform the least accurate in most conditions and never outperforms the other methods, but the differences in accuracy become less pronounced with larger sample and effect sizes.

@@

To assess the effect of the treatment-effect difference $\beta$, in addition to the \textit{General simulation design}, the following facet was added in this study: 
\begin{enumerate}
\item[7.] Two levels for $\beta$, the global mean difference in treatment outcomes: $\beta = 2.5$ (corresponding to a medium effect size, Cohen's $d = 0.5$) and $\beta = 5.0$ (corresponding to a large effect size; Cohen's $d = 1.0$). 
\end{enumerate}

@@

The interactions in Study~I (Figure~\ref{fig:modelC}) can be referred to as piecewise interactions, as their effect is a stepwise function of the moderator (partitioning) variables. Trees are preeminently suited for recovering such piecewise or subgroup interactions, but may have difficulty when the true interactions are continuous functions of moderator variables (for example, $U_1 \cdot U_2$). At the same time, linear regression models with pre-specified interaction terms may perform well in recovering continuous interactions, but may have difficulty in recovering piecewise interactions. Therefore, in the third simulation study, in addition to the \textit{General simulation design} described above, the following facet was added:

@@

For generating datasets with both piecewise and continuous interactions, the partition as depicted in Figure~\ref{fig:modelC} was also used. In addition, the fixed-effects part $x_{i}^{\top}\beta_{j}$ in each of the terminal nodes now comprised continuous main and (treatment) interaction effects of the partitioning variables. In other words, the partitioning variables $U_2$, $U_1$ and $U_5$ also appear in the linear predictor $x_i$, as part of the terms presented in Table~\ref{tab:continuous_terms}. The corresponding node-specific $\beta_j$ parameters are also presented in Table~\ref{tab:continuous_terms}. The $\beta_j$ values were chosen to yield the same treatment-subgroup means as in Figure~\ref{fig:modelC}. The interaction terms were created using centered $U_k$ variables, calculated by subtracting their variable means. Again, the outcome variable was calculated as $y_i = x_{i}^{\top}\beta_{j} + z_{i}^{\top}b + \epsilon_i$.

In datasets with purely continuous interactions, $\beta$ has a global value and no subscript, comprising only purely continuous main and interaction effects, as shown by terms and the single column for $\beta$ in Table~\ref{tab:continuous_terms}. The outcome variable was calculated as $y_i = x_{i}^{\top}\beta + z_{i}^{\top}b + \epsilon_i$.

@@

As the outcome variable is continuous, we employed an identity link and Gaussian response distribution. The resulting GLMM trees will therefore be refered to as LMM trees. To compare the accuracy of LMM trees, we also fitted LM trees and LMMs with pre-specified interactions to the data. In the LMMs, the outcome variable was regressed on a random intercept, main effects of treatment and the potential moderators (partitioning variables) and interactions between treatment and the potential moderators. As it is not known in advance how to interact the potential moderators, higher-order interactions were not included.

@@

Figure~\ref{fig:lmtree_C&W} and Figure~\ref{fig:lmertree_C&W} present the LM and LMM trees fitted on the IPDMA data. The LM tree (Figure~\ref{fig:lmtree_C&W}) selected level of education as the first partitioning variable, and presence of a comorbid anxiety disorder as a second partitioning variable, for observations with a higher level of education. By taking into account study-specific intercepts, the LMM tree (Figure~\ref{fig:lmertree_C&W}) indicates that the first split in the LM tree may be spurious. The LMM tree selected presence of a comorbid anxiety disorder as the only partitioning variable. The terminal nodes of Figure~\ref{fig:lmertree_C&W} show only a single treatment-subgroup interaction: for patients without a comorbid anxiety disorder, CBT and PHA provide more or less the same reduction in HAM-D scores (Cohen's $d = 0.05$). For patients with a comorbid anxiety disorder, PHA provides a greater reduction in HAM-D scores (Cohen's $d = 0.39$). The estimated intraclass correlation coefficient for the LMM tree was .05.

@@

The LMM with pre-specified treatment interactions yielded three significant predictors of treatment outcome: like in the LMM tree, an effect of the presence of a comorbid anxiety disorder was found (main effect: $b = 2.29$, $p = .002$; interaction with treatment: $b = -2.10$, $p = .028$). Also, the LMM indicated an interaction between treatment and age ($b = .10$, $p = .018$).

@@

We presented the GLMM tree algorithm, which allows for estimation of a GLM-based recursive partition, as well as estimation of global random-effects parameters. We hypothesized GLMM trees to be well suited for the detection of treatment-subgroup interactions in clustered datasets. We confirmed this through our simulation studies and by applying the algorithm to an existing dataset on the effects of depression treatments.

Our simulations focused on the performance of the GLMM tree algorithm in datasets with continuous response variables. The resulting LMM trees accurately recovered the subgroups in 90\% of simulated datasets with treatment-subgroup interactions, outperforming LM trees without random effects. In terms of predictive accuracy, LMM trees outperformed LM trees as well as MERTs, predicting treatment-effect differences in test data with .94 accuracy, on average. In datasets without treatment-subgroup effects, LMM trees showed a rather low Type-I error rate of 4\%, compared with a Type-I error rate of 33\% for LM trees.

Better performance of LMM trees was mostly observed when random effects were sizeable and correlated with potential partitioning variables. In those circumstances, LM trees are likely to detect spurious splits and subgroups. Especially with smaller effect and sample sizes (i.e., Cohen's $d=.5$ and/or $N = 200$), LMM trees outperformed the other tree methods.  As such effect and sample sizes are quite common in multi-center clinical trials, the GLMM tree algorithm may provide a useful tool for subgroup detection in those instances.

In the absence of random effects, LM and LMM trees yielded very similar predictive accuracy. This finding is of practical importance, indicating that application of the GLMM tree algorithm will not reduce accuracy when in fact, random effects are absent from the data.

Compared to LMMs with pre-specified interactions, LMM trees provided somewhat better accuracy, on average. LMM trees performed much better than LMMs when interactions were at least partly piecewise. However, the performance ofLMM trees deteriorated when the interactions were purely continuous functions of the predictor variables, in which case LMMs with pre-specified interactions performed very well. The performance of LMMs deteriorated with a larger number of predictor variables, whereas the performance of LMM trees was not affected by this, confirming our expectation that LMM trees are better suited for exploration than LMMs.

In the Application, we found the LMM tree yielded somewhat higher predictive accuracy, while using a smaller number of variables than the LM tree and the LMM with pre-specified interactions. The LMM trees obtained over repeated subsamples of the training data proved to be relatively stable.

\subsubsection{Limitations and future directions}

Recursive partitioning methods were originally developed as a non-parametric tool for classification and regression, assuming the mechanism that generated the data unknown \cite<e.g., >{Brei01TwoCult}. The GLMM tree algorithm takes MOB trees and GLMMs as building blocks and thereby inherits some of their sensitivity to model misspecifications.

As with GLMMs, misspecification of the random effects may negatively affect the accuracy of GLMM trees. Previous research on GLMMs has shown that misspecifying the shape of the random-effects distribution reduces the accuracy of random-effects predictions, but has little effects on the fixed-effects parameter estimates \cite<e.g., >{NeuhyMcCu12}. This indicates that for GLMM trees, misspecification of the shape of the random-effects distribution will affect random-effects predictions, but will have little effect on the estimated tree. The incorrect specification of a predictor as having a fixed instead of a random effect may be more likely to result in spurious splits or poorer parameter estimates. Therefore, to check for model misspecification, users are advised to visually inspect residuals and predicted random effects. The tutorial included in the supplementary material shows how residuals and random effects predictions can be obtained and plotted.

As with MOB trees, GLMM trees will perform best when the true data structure is tree-shaped. Trees may need many splits to approximate structures with a different shape, which may be more unstable under small perturbations. Therefore, in the Application we have shown how tree stability can be assessed using the methods developed by \citeA{PhilyZeil16}. Furthermore, when relevant partitioning variables are omitted, GLMM tree can only approximate the subgroups using the specified variables. Although our simulations did not involve scenarios where relevant partitioning variables were omitted, the results of \citeA{FricyStro14} indicate that MOB trees can detect parameter instability even when covariates are only loosely connected to the group structure.

GLMM trees rely more on user specification of predictor variables than methods like CART and MERT. On the one hand, this increases the danger of model misspecification. On the other hand, it increases power and accuracy if the fixed-effects predictor variable(s) are correctly specified. For example, the best approach for assessing treatment-effect differences with MERTs is not completely obvious. We tried two approaches in our simulations: firstly, fitting a single MERT with the treatment indicator as one of the potential partitioning variables and secondly, fitting separate MERTs for each level of the treatment indicator. With the first approach, MERTs often did not detect the treatment indicator as a predictor variable in datasets with treatment-subgroup interactions. The second approach yielded better accuracy, indicating that MERTs profit from user specification of the relevant predictor variable(s). The GLMM tree algorithm provides a straightforward approach for including relevant fixed-effects predictor variables in the model. Our results suggest this yields higher predictive accuracy if the model is correctly specified. However, our simulations have not assessed the effect of misspecification of the fixed-effects predictor variable(s) and such misspecification may reduce predictive accuracy.

For recursive partitioning methods fitting local parametric models only, the danger of model misspecification may be limited \cite<e.g., >{Ciam91, SiciyMass08, SuyWang04}. GLMM trees, however, fit a global random-effect model in addition to local fixed-effects regression models and may therefore be less robust against model misspecification. Further research on the effects of model misspecifications on the performance of GLMM trees is therefore needed.

In the Introduction, we mentioned several existing tree-based methods for treatment-subgroup interaction detection. These methods have different objectives and there is not yet an agreed-upon single best method. In a simulation study, \citeA{SiesyVanM16} found the method of \citeA{ZhanyTsia12Stat} to perform best, followed by MOB. However, the method of Zhang et al. performed worst under some conditions of the simulation study in terms of the Type I error rate. Further research comparing tree-based methods for treatment-subgroup interaction detection is needed, especially  for clustered datasets, as our simulations and comparisons mostly focused on LMM trees and LM trees.
